Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) \& The State of Sustainable Software -- TITLE NEEDS WORK

### 2018-03-28 WSSSPE5.1_report/WSSSPE5.1_paper.tex
adding start of WSSSPE5.1 report

\title[]{Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}

\author{Daniel~S.~Katz$^{(1)}$,
Kyle~E.~Niemeyer$^{(2)}$,
Sandra~Gesing$^{(3)}$,
Lorraine~Hwang$^{(4)}$,
Wolfgang~Bangerth$^{(5)}$,
Simon~Hettrick$^{(6)}$,
Ray Idaszak$^{(7)}$,
Jean~Salac$^{(8)}$,
Neil~Chue~Hong$^{(9)}$,
Santiago~N\'u\~nez-Corrales$^{(10)}$,
Alice~Allen$^{(11)}$,
R.~Stuart~Geiger$^{(12)}$,
Jonah~Miller$^{(13)}$,
Emily~Chen$^{(14)}$,
Anshu~Dubey$^{(15)}$,
Patricia~Lago$^{(16)}$
%\\Author order to be determined based on contributions
}

%
\thanks{{}$^{(1)}$ National Center for Supercomputing Applications (NCSA) \&
Department of Computer Science  \&
Department of Electrical and Computer Engineering  \&
School of Information Sciences (iSchool),
University of Illinois, Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
%
\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
%
\thanks{{}$^{(3)}$ Center for Research Computing \& Department of Computer Science and Engineering,
University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}
%
\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: 0000-0002-1021-3101}
%
\thanks{{}$^{(5)}$ Department of Mathematics, Colorado State
  University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
%
\thanks{{}$^{(6)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
%
\thanks{{}$^{(7)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: 0000-0002-3444-7615}
%
\thanks{{}$^{(8)}$ University of Chicago, Chicago, IL, USA; salac@uchicago.edu; ORCID: 0000-0001-5971-9333}
%
\thanks{{}$^{(9)}$ University of Edinburgh, UK; N.ChueHong@epcc.ed.ac.uk;
ORCID: 0000-0002-8876-7606}
%
\thanks{{}$^{(10)}$ Illinois Informatics Institute (I3) and National Center for Supercomputing Applications (NCSA), University of Illinois at Urbana-Champaign, IL, USA;
nunezco2@illinois.edu;
ORCID: 0000-0003-4342-6223}
%
\thanks{{}$^{(11)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
%
\thanks{{}$^{(12)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
%
\thanks{{}$^{(13)}$ Perimeter Institute for Theoretical Physics \&
Department of Physics, University of Guelph,
Guelph, ON, Canada;
jmiller@perimeterinstitute.ca; ORCID: 0000-0001-6432-7860}
%
\thanks{{}$^{(14)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
%
\thanks{{}$^{(15)}$ Mathematics and Computer Science Division,
Argonne National Laboratory,
Lemont, IL, USA \&
University of Chicago,
Chicago, IL, USA;
Adubey@anl.gov;
ORCID: 0000-0003-3299-7426}
%
\thanks{{}$^{(16)}$ Vrije Universiteit Amsterdam, The Netherlands; p.lago@vu.nl; ORCID: 0000-0002-2234-0845}

\begin{abstract}

This article summarizes motivations, organization, and activities of the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4). The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit. This article discusses the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices. The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group. Finally, it discusses a survey of the workshop attendees.

@@

The Fourth Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
held over 2 1/2 days on 12--14 September 2016 in Manchester, England, with 68 attendees.
This location and date was selected so that WSSSPE4 immediately preceded the
First Research Software Engineers (RSE) Conference.
Previous WSSSPE events\footnote{The first WSSSPE workshop was named
``Working towards
Sustainable Software for Science: Practice and Experiences,'' which remains the meaning
of the WSSSPE group, but the workshops after that were named
``Workshop on Sustainable
Software for Science: Practice and Experiences.'' Together these reflect
that WSSSPE is both a community and a set of workshops.}, all of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1, WSSSPE2-pre-report,WSSSPE2},
two half-day workshops\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}, \url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}} held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, and a 1-and-1/2-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}.
%
Based on the work done at these previous WSSSPE meetings, WSSSPE4 was aimed
at producing working groups that would better continue working after the workshop ended.

@@

This paper is a summary of a longer report~\cite{WSSSPE4-report} on the workshop and the submitted materials.
The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).  One full day of the
workshop was spent with participants in the working groups, which occurred in parallel
with each other.  And each of the working groups left with a plan for how they could move
forward.
This report also highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
%An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Call for participation} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Post-WSSSPE3 feedback said that attendees
had two distinct motivations for attending:  to make a better future
for research software, and to immediately do better research software development. Thus,
WSSSPE4 was %partially 
planned as two tracks:

\begin{quote}
    \textbf{Track 1 -- Building a sustainable future for open-use research
    software}: defining a vision of the future of open-use
    research software, and in the workshop, initiating the activities that are
    needed to get there.

    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
    improving the quality of today's research software and the
    experiences of its developers by sharing best practices and experiences.
\end{quote}

The call for participation requested:
idea papers, implementable proposals related to Track~1;
position papers, longer, not previously published papers
discussing what we can do to improve sustainable scientific
software in the short term;
experience papers, longer papers that discuss current
experiences and how they have been used to improve the quality of
today's research software and/or the experiences of its developers;
demos, brief papers describing a tool or
process that would be demonstrated that improves the quality of today's research
software and\slash or the experiences of its developers; and
extended abstracts describing lightning talks,
where each author could make a brief statement about work that either had been
done or was needed.

@@

Submissions to WSSSPE4 comprised
19 lightning talks,
4 idea papers,
3 position paper,
5 experience papers,
and
3 demos.
Most submissions were accepted, since the goal of WSSSPE is always to
take in as many relevant inputs as possible, and to encourage their authors to
participate in sharing and implementing their ideas. Two submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
were accepted as lightning talks.
% Specifically,
% 19 lightning talks,
% 4 idea papers,
% 3 position papers,
% 2 experience papers,
% and
% 3 demos
% were accepted, as mentioned in Section~\ref{sec:presentations}.
%
The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.

@@

WSSSPE4 introduced a Code of Conduct (CoC)\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}, which was conceived for the workshop and as guidance for the community of scientists that WSSSPE
supports, including their personal and online interactions (e.g., on
Twitter, in email lists, in Slack). This CoC is based on the
FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
CoC~\cite{Code4Lib:CoC}.
In introducing the CoC, we asked participants to agree to the following main guidelines:
\begin{quote}
    WSSSPE events are community events intended for networking and collaboration
    as well as learning. We value the participation of every member of the
    community and want all attendees to have an enjoyable and fulfilling
    experience. Accordingly, all attendees are expected to show respect and
    courtesy to other attendees throughout the event and in interactions online
    associated with the event.

@@

%     \begin{itemize}
%     \item Communicate appropriately for a professional audience including
%     people of many different backgrounds. Sexual language and imagery are not
%     appropriate for any event.

%     \item Be kind to others. Do not insult or put down other attendees. Be
%     mindful of jargon, which can sometimes exclude others from engaging in the
%     discussion.

%     \item Behave professionally. Remember that harassment and sexist, racist,
%     ageist, or exclusionary behavior are not appropriate.
%     \end{itemize}
 \end{quote}

The CoC was discussed at the beginning of WSSSPE4, with the CoC subcommittee
and a general email address for reporting concerns or incidents, or
asking questions.  One concern was raised after the first half day, 
% regarding how part of the workshop was being run, 
and we changed the workshop
to address this concern.

@@

%\note{by Lorraine and Dan}

Going into WSSSPE4, WSSSPE had no formal mission or vision statement.
The organizers developed a strawhorse, which was presented to the participants early in the meeting.
The presentation included guidelines as well as examples from other similar communities.
The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world would look like if the organization is successful; and from which, focus areas could be used to establish the scope of the organization along with its goals.

The participants were 
%given time after the presentation to write down their  
invited to comment,
% Seven people volunteered to work on 
and 
%revising 
the mission and vision statements, were revised 
%incorporating these comments and presenting back to the community.
based on this feedback.
% the committee redrafted the mission and vision.
The committee added focus areas to clarify the organization's breadth and to keep the mission and vision simple and long-lasting.
Comments after the close of the meeting were also incorporated into the statements.
A final draft was put on GitHub and advertised
via the WSSSPE mailing list, Facebook group, and Twitter.
After two weeks without suggested changes, the final statements\footnote{\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}}
%as shown below, 
were added
to the WSSSPE website.

% {\bf Mission.}
% WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.

% {\bf Vision.}
% We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

% {\bf Focus areas.}
% WSSSPE promotes sustainable research software by positively impacting:
% \begin{itemize}
% \item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
% \item {\bf Careers}. Developing and supporting career paths in research software development and engineering
% \item {\bf Learning}. Engaging in activities to promote peer learning and interaction
% \item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
% \end{itemize}

% \textbf{Definitions:}
% \emph{Sustainable software} has the capacity to endure such that it will continue to
% be available in the future, on new platforms, meeting new needs.
% The \emph{research software lifecycle} includes:
% acquiring and assembling resources (including funding and people) into teams and communities,
% developing software,
% using software,
% recognizing contributions to and of software,
% and
% maintaining software.

@@

%\note{from Patricia}

The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
%
Sustainability is broadly associated with natural ecologic systems. When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}. In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.  Lago's keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability, providing a complementary angle to the workshop discussions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Papers and demos} \label{sec:papers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\note{Sandra Gesing will write this}

WSSSPE4 included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
two experience papers, and three demos) that addressed a wide range of topics around
sustainability for software in science. These talks covered three main areas.
%
% The following four papers belong to the first area: the academic environment.
%For talks with multiple authors, an
% \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
%Four papers %,  
% \begin{itemize}
% \item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
% Improving Incentives for Quality Software}
%~\cite{Heroux:2016ws}, 
% \item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
% Society for Research Software}
%~\cite{GAllen:2016ws}, 
% \item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
% \emph{Experience Paper: Preliminary analysis of a survey of UK Research
% Software Engineers}
%~\cite{Philippe:2016ws}, and 
% \item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
% create and develop research ties within academia}
%~\cite{Gwilliams:2016ws} 
%belong to 
The first area was the academic environment, with four papers
that address diverse aspects ranging from incentives for quality software to
advocating a professional society for research software to roles and degrees for
research software engineers~\cite{Heroux:2016ws, GAllen:2016ws, Philippe:2016ws, Gwilliams:2016ws}.
%\end{itemize}
The second area was concerned with characteristics and needs of
research software and included five papers~\cite{Dubey1:2016ws, ChueHong:2016ws, Dubey2:2016ws, Queiroz:2016ws, Childers:2016ws}.
% \begin{itemize}
% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
% Engineering and Community Codes Track in ATPESC}
% \item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
% research software, and how should we do it?}
% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
% of Software for Scientific Simulation}
% \item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
% Collaborative Gamification Design for Scientific Software}
% \item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
% David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
% the Way Digital Libraries will Operate}
% \end{itemize}
%\noindent 
%Three contributions, ~\cite{Ganguly:2016ws}, ~\cite{Shende:2016ws},
%and ~\cite{Sallai:2016ws} , focused on 
The third  area focused on elucidating lessons learned
from or visions for use cases of scientific software and communities working
with and\slash or on a software package, with three papers~\cite{Ganguly:2016ws, Shende:2016ws, Sallai:2016ws}.

In addition, 19 lightning talks were presented at WSSSPE4~\cite{Sufi:2016ws, Gesing:2016ws, Ram:2016ws, Loffler:2016ws, Katz:2016ws, Idaszak:2016ws, Hwang:2016ws, Goble:2016ws, Druskat:2016ws, Contrastin:2016ws, AAllen:2016ws, ChueHong:2016wsb, vanHage:2016ws, GAllen:2016wsb, Seidel:2016ws, Emsley:2016ws, Dongarra:2016ws, Bauer:2016ws, Aldabjan:2016ws}.

% \begin{itemize}
% \item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
% Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
% Reproducing, and Sharing Simulations}
% \item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
% Performance Evaluation of Scientific Software}
% \item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
% \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
% Atom-Typing for Molecular Simulation}
% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Lightning talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe4/agenda/}{Slides.}}
\end{comment}

%
%For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.

% \begin{itemize}%[itemsep=1ex]
%     \item Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
%     programme; supporting the social side of research software}

%     \item Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
%     Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
%     \emph{The Science Gateways Community Institute - Supporting Communities to
%     Achieve Sustainability for Their Science Gateways}

%     \item Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
%     and Scott Chamberlain: \emph{A model for
%     peer review and onboarding research software}
%     \item Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
%     \emph{A vision of computing in 10+ years}
%     \item Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
%     Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
%     \emph{Software Citation: Process, Principles, and
%       Implementation}

%     \item Ray Idaszak\textsuperscript{\textasteriskcentered},
%     David G.~Tarboton, Hong Yi, Michael Stealey,
%     Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
%     Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
%     Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
%     Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
%     Best Practices and Culture Change for Developing Sustainable Community
%     Software}

%     \item Lorraine Hwang\textsuperscript{\textasteriskcentered},
%     Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
%     \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
%     Community}

%     \item Carole Goble: \emph{A Simple Profiling Framework for Software
%     User-Producer Reciprocity Review}
%     \item Stephan Druskat: \emph{A proposal for the measurement and
%     documentation of research software sustainability in interactive metadata
%     repositories}

>>>>

\title[]{Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) \& The State of Sustainable Software -- TITLE NEEDS WORK}

\author{Daniel~S.~Katz$^{(1)}$,
Caroline $^{(2)}$,
Rob $^{(3)}$,
Alexander $^{(4)}$,
Stephan\ $^{(5)}$\
please add full names, affiliations, orcids
%Kyle~E.~Niemeyer$^{(2)}$,
%Sandra~Gesing$^{(3)}$,
%Lorraine~Hwang$^{(4)}$,
%Wolfgang~Bangerth$^{(5)}$,
%Simon~Hettrick$^{(6)}$,
%Ray Idaszak$^{(7)}$,
%Jean~Salac$^{(8)}$,
%Neil~Chue~Hong$^{(9)}$,
%Santiago~N\'u\~nez-Corrales$^{(10)}$,
%Alice~Allen$^{(11)}$,
%R.~Stuart~Geiger$^{(12)}$,
%Jonah~Miller$^{(13)}$,
%Emily~Chen$^{(14)}$,
%Anshu~Dubey$^{(15)}$,
%Patricia~Lago$^{(16)}$
%\\Author order to be determined based on contributions
}

%
\thanks{{}$^{(1)}$ National Center for Supercomputing Applications (NCSA) \&
Department of Computer Science  \&
Department of Electrical and Computer Engineering  \&
School of Information Sciences (iSchool),
University of Illinois, Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
%
\thanks{{}$^{(2)}$ department/school, city, country; email; ORCID: orcid}
%
\thanks{{}$^{(3)}$ department/school, city, country; email; ORCID: orcid}
%
\thanks{{}$^{(4)}$ department/school, city, country; email; ORCID: orcid}
%
\thanks{{}$^{(5)}$ department/school, city, country; email; ORCID: orcid}
%

\begin{abstract}
\todo{Need to update}
This article summarizes motivations, organization, and activities of the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4). The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit. This article discusses the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices. The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group. Finally, it discusses a survey of the workshop attendees.

@@

\todo{Dan}

The Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE5.1)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe5/}} \katznote{check URL} was a one-day meeting held in September 2017 in Manchester, England immediately preceding the
Second Research Software Engineers (RSE) Conference, so that RSE attendees could also attend WSSSPE5.1.  \katznote{number of attendees?}

Previous WSSSPE events\footnote{The first WSSSPE workshop was named
``Working towards
Sustainable Software for Science: Practice and Experiences,'' which remains the meaning
of the WSSSPE group, but the workshops after that were named
``Workshop on Sustainable
Software for Science: Practice and Experiences.'' Together these reflect
that WSSSPE is both a community and a set of workshops.}, all but the last of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1, WSSSPE2-pre-report,WSSSPE2},
two half-day workshops\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}, \url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}} held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, a 1-and-1/2-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}, and a 2-and-1/2-day workshop that immediately preceded the
First Research Software Engineers (RSE) Conference in Manchester, England.
~\cite{WSSSPE4}
%
Based on the work done at these previous WSSSPE meetings, WSSSPE4 was aimed
at producing working groups that would better continue working after the workshop ended.

@@

The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).  One full day of the
workshop was spent with participants in the working groups, which occurred in parallel
with each other.  And each of the working groups left with a plan for how they could move
forward.
This report also highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).

\katznote{new text below, need to replace lots of what is above:}

In October 2017, a group of people interested in sustainable research software came together at the Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) meeting in Manchester, UK. This workshop built on six previous WSSSPE workshops, all of which dedicated significant portions of the meeting to group discussions about problems and potential solutions in the sustainable research software space.  WSSSPE5.1 used the speed blog methodology to generate 8 reports on different views of the space. These blogs, previously published by the UK Software Sustainability Institute (SSI) on its website, form the second portion of this paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Call for participation and response} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Stephan, maybe you can put something short here?}

%Post-WSSSPE3 feedback said that attendees
%had two distinct motivations for attending:  to make a better future
%for research software, and to immediately do better research software development. Thus,
%WSSSPE4 was planned as two tracks:
%
%\begin{quote}
%    \textbf{Track 1 -- Building a sustainable future for open-use research
%    software}: defining a vision of the future of open-use
%    research software, and in the workshop, initiating the activities that are
%    needed to get there.
%
%    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
%    improving the quality of today's research software and the
%    experiences of its developers by sharing best practices and experiences.
%\end{quote}
%
%The call for participation requested:
%idea papers, implementable proposals related to Track~1;
%position papers, longer, not previously published papers
%discussing what we can do to improve sustainable scientific
%software in the short term;
%experience papers, longer papers that discuss current
%experiences and how they have been used to improve the quality of
%today's research software and/or the experiences of its developers;
%demos, brief papers describing a tool or
%process that would be demonstrated that improves the quality of today's research
%software and\slash or the experiences of its developers; and
%extended abstracts describing lightning talks,
%where each author could make a brief statement about work that either had been
%done or was needed.

@@

Submissions to WSSSPE4 comprised
19 lightning talks,
4 idea papers,
3 position paper,
5 experience papers,
and
3 demos.
Most submissions were accepted, since the goal of WSSSPE is always to
take in as many relevant inputs as possible, and to encourage their authors to
participate in sharing and implementing their ideas. Two submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
were accepted as lightning talks.
The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.

@@

\todo{do we want to keep some discussion of this here?  maybe fold in into the intro?}

WSSSPE5.1 included a Code of Conduct (CoC)\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}} as guidance for the community of scientists that WSSSPE
supports, including the workshop and their personal and online interactions (e.g., on
Twitter, in email lists, in Slack). This CoC is based on the
FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
CoC~\cite{Code4Lib:CoC}.
In introducing the CoC, we asked participants to agree to the following main guidelines:
\begin{quote}
    WSSSPE events are community events intended for networking and collaboration
    as well as learning. We value the participation of every member of the
    community and want all attendees to have an enjoyable and fulfilling
    experience. Accordingly, all attendees are expected to show respect and
    courtesy to other attendees throughout the event and in interactions online
    associated with the event.

@@

 \end{quote}

The CoC was discussed at the beginning of WSSSPE5.1, with the CoC subcommittee
and a general email address for reporting concerns or incidents, or
asking questions.

@@

\todo{should we take any of this into the intro, or drop it all?}

 {\bf Mission.}
 WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.

 {\bf Vision.}
 We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

 {\bf Focus areas.}
 WSSSPE promotes sustainable research software by positively impacting:
 \begin{itemize}
 \item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
 \item {\bf Careers}. Developing and supporting career paths in research software development and engineering
 \item {\bf Learning}. Engaging in activities to promote peer learning and interaction
 \item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
 \end{itemize}

 \textbf{Definitions:}
 \emph{Sustainable software} has the capacity to endure such that it will continue to
 be available in the future, on new platforms, meeting new needs.
 The \emph{research software lifecycle} includes:
 acquiring and assembling resources (including funding and people) into teams and communities,
 developing software,
 using software,
 recognizing contributions to and of software,
 and
 maintaining software.

@@

\todo{Stephan}

%The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
%Sustainability is broadly associated with natural ecologic systems. When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}. In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.  Lago's keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability, providing a complementary angle to the workshop discussions.
%
%WSSSPE4 included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
%two experience papers, and three demos) that addressed a wide range of topics around
%sustainability for software in science. These talks covered three main areas.
%The first area was the academic environment, with four papers
%that address diverse aspects ranging from incentives for quality software to
%advocating a professional society for research software to roles and degrees for
%research software engineers~\cite{Heroux:2016ws, GAllen:2016ws, Philippe:2016ws, Gwilliams:2016ws}.
%The second area was concerned with characteristics and needs of
%research software and included five papers~\cite{Dubey1:2016ws, ChueHong:2016ws, Dubey2:2016ws, Queiroz:2016ws, Childers:2016ws}.
%The third  area focused on elucidating lessons learned
%from or visions for use cases of scientific software and communities working
%with and\slash or on a software package, with three papers~\cite{Ganguly:2016ws, Shende:2016ws, Sallai:2016ws}.
%
%In addition, 19 lightning talks were presented at WSSSPE4~\cite{Sufi:2016ws, Gesing:2016ws, Ram:2016ws, Loffler:2016ws, Katz:2016ws, Idaszak:2016ws, Hwang:2016ws, Goble:2016ws, Druskat:2016ws, Contrastin:2016ws, AAllen:2016ws, ChueHong:2016wsb, vanHage:2016ws, GAllen:2016wsb, Seidel:2016ws, Emsley:2016ws, Dongarra:2016ws, Bauer:2016ws, Aldabjan:2016ws}.
%
%The final WSSSPE4 presentation was a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
%These were: Alice Allen, Editor, Astrophysics Source Code Library, who brought an understanding of the difficulties of organizing a community and curating their software;
%Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
%Patricia Lago from Vrije Universiteit Amsterdam, who brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
%Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Speed blogs} \label{sec:speed_blogs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Alexander}

\note{explain the process of speed blogging}

\note{summarize the speed blogs themselves, with citations/pointers to them on the SSI page}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analyzing the speed blogs} \label{sec:speed_blog_analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Caroline and Rob}

\note{explain the method?}

\note{show the date?}

\note{show a figure?}

\note{explain the results?}

\katznote{here's some stuff I wrote previously - see if you want to keep any of it:}

However, to frame the context of these reports, we first examine the overall space, which can be viewed in many ways.  The high-level topics in the call for participation in the previous meeting, WSSSPE4, were: development and community, professionalization, training, credit, software publishing, software discoverability and reuse, and reproducibility and testing. The set of topics initially proposed in WSSSPE5.1 were: social factors, technical choices, funding and careers, management and guidance, citation and credit, training, reproducible research, software quality, legacy code, and containers and packaging. WSSSPE5.1 also mentioned that there seemed to be sufficient writing already on citation and credit, and containers and packaging, so while these topics that fit under the WSSSPE umbrella, they were not suggested as topics for speed blogs. In addition, the lead author of this paper has given talks to summarize this space previously, based on topics that emerged from the previous WSSSPE meetings before WSSSPE4. Those topics are: software engineering; portability; training and education; incentives, citation/credit models, and metrics; intellectual property; publication and peer review; software communities and sociology; sustainability and funding models; career paths; software dissemination, catalogs, search, and review; multi-disciplinary science; and reproducibility. All three of these ways of looking at the space have value; none are wrong, none are right, all are useful. They each try to look at the problems we have.

Another choice is to look at this at a different level: that of actors. We can say that we are concerned about software, people, organizations, and the field itself. And we could center our view through the lens of one or more of these actors.
